Examining background gene lists using simulation - comparison of different enrichment techniques

### 2024-06-13 analysis/sim/run_simpw.Rmd
fig3
sims
new sims

11. clusterprofiler_fixed up gene sets

12. clusterprofiler_fixed down gene sets

13. clusterprofiler_nobg up gene sets

14. clusterprofiler_nobg down gene sets

15. fora up gene sets

16. fora down gene sets

17. fgsea up gene sets

18. fgsea down gene sets

19. mitch up

20. mitch dn

@@

SIMS=500
#run 500 sims
FRAC_DE=0.05
FC=1
N_REPS=3
DGE_FUNC="deseq2"
SUM_COUNT=2e7
VARIANCE=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1,1.1,1.2,1.3,1.4,1.5)

@@

## Plotfig1_ORAjac.png

@@

cp <- subset(res,PWAY_FUNC == "clusterprofiler")
fo <- subset(res,PWAY_FUNC == "fora")
fg <- subset(res,PWAY_FUNC == "fgsea")

cp %>% kbl(caption="clusterprofiler") %>% kable_paper("hover", full_width = F)
fo %>% kbl(caption="fora") %>% kable_paper("hover", full_width = F)
fg %>% kbl(caption="fg") %>% kable_paper("hover", full_width = F)

@@

plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")

legend("bottomleft", inset=.02, title="tool",
   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
   horiz=FALSE, cex=1.2, pch=19,lwd=2)

plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")

plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")

png("fig3_sim.png",width=720, height=480)
par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
par(mfrow=c(1,3))

plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")

legend("bottomleft", inset=.02, title="tool",
   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
   horiz=FALSE, cex=1.2, pch=19,lwd=2)

plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")

plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")

dev.off()

>>>>

11. clusterprofiler bg fix up gene sets

12. clusterprofiler bg fix down gene sets

13. clusterprofiler FDR fix up gene sets

14. clusterprofiler FDR fix down gene sets

15. clusterprofiler BG and FDR fix up gene sets

16. clusterprofiler BG and FDR fix down gene sets

17. fora up gene sets

18. fora down gene sets

19. fgsea up gene sets

20. fgsea down gene sets

@@

SIMS=1000
#run 500 sims
FRAC_DE=0.05
FC=0.5
N_REPS=3
DGE_FUNC="deseq2"
SUM_COUNT=2e7
VARIANCE=c(0,0.1,0.2,0.3,0.4,0.5,0.6)

@@

## Plot

fig1_ORAjac.png

@@

cp <- subset(res,PWAY_FUNC == "clusterProfiler default")
cpbg <- subset(res,PWAY_FUNC == "clusterProfiler BG fix")
cpfdr <- subset(res,PWAY_FUNC == "clusterProfiler FDR fix")
cpbgfdr <- subset(res,PWAY_FUNC == "clusterProfiler BG and FDR fix")
fo <- subset(res,PWAY_FUNC == "fora")
fg <- subset(res,PWAY_FUNC == "fgsea")

cp %>% kbl(caption="clusterProfiler") %>% kable_paper("hover", full_width = F)
cpbg %>% kbl(caption="clusterProfiler with background bug fix") %>% kable_paper("hover", full_width = F)
cpfdr %>% kbl(caption="clusterProfiler with FDR bug fix") %>% kable_paper("hover", full_width = F)
cpbgfdr %>% kbl(caption="clusterProfiler with background and FDR fixes") %>% kable_paper("hover", full_width = F)
fo %>% kbl(caption="fora") %>% kable_paper("hover", full_width = F)
fg %>% kbl(caption="fg") %>% kable_paper("hover", full_width = F)

@@

plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
points(cpbg$VARIANCE,cpbg$p,type="b",pch=19,col="orange")
points(cpfdr$VARIANCE,cpfdr$p,type="b",pch=19,col="darkgreen")
points(cpbgfdr$VARIANCE,cpbgfdr$p,type="b",pch=19,col="purple")
points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")

legend("bottomleft", inset=.02, title="tool",
   c("CP default","CP BG fix","CP FDR fix","CP BG and FDR fix","fora","fgsea"),
   col=c("black","orange","darkgreen","purple","red","blue"),horiz=FALSE, cex=1.2, pch=19,lwd=2)

plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
points(cpbg$VARIANCE,cpbg$r,type="b",pch=19,col="orange")
points(cpfdr$VARIANCE,cpfdr$r,type="b",pch=19,col="darkgreen")
points(cpbgfdr$VARIANCE,cpbgfdr$r,type="b",pch=19,col="purple")
points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")

plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
points(cpbg$VARIANCE,cpbg$f,type="b",pch=19,col="orange")
points(cpfdr$VARIANCE,cpfdr$f,type="b",pch=19,col="darkgreen")
points(cpbgfdr$VARIANCE,cpbgfdr$f,type="b",pch=19,col="purple")
points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")

png("fig3_sim.png", width=7,height=5,units="in",res=150,pointsize=12)
par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
par(mfrow=c(1,3))

plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
points(cpbg$VARIANCE,cpbg$p,type="b",pch=19,col="orange")
points(cpfdr$VARIANCE,cpfdr$p,type="b",pch=19,col="darkgreen")
points(cpbgfdr$VARIANCE,cpbgfdr$p,type="b",pch=19,col="purple")
points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
grid()
legend("bottomleft", inset=.02, title="tool",
   c("CP default","CP BG fix","CP FDR fix","CP BG and FDR fix","fora","fgsea"),
   col=c("black","orange","darkgreen","purple","red","blue"),
   horiz=FALSE, cex=1.2, pch=19,lwd=2)

plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
points(cpbg$VARIANCE,cpbg$r,type="b",pch=19,col="orange")
points(cpfdr$VARIANCE,cpfdr$r,type="b",pch=19,col="darkgreen")
points(cpbgfdr$VARIANCE,cpbgfdr$r,type="b",pch=19,col="purple")
points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
grid()

plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
points(cpbg$VARIANCE,cpbg$f,type="b",pch=19,col="orange")
points(cpfdr$VARIANCE,cpfdr$f,type="b",pch=19,col="darkgreen")
points(cpbgfdr$VARIANCE,cpbgfdr$f,type="b",pch=19,col="purple")
points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
grid()
dev.off()

Barplot is simpler.

```{r,bars}

prec <- c("CP default"=mean(cp$p),"CP BG fix"=mean(cpbg$p),"CP FDR fix"=mean(cpfdr$p),"CP BG and FDR fix"=mean(cpbgfdr$p),"fora"=mean(fo$p),"fgsea"=mean(fg$p))
rec <- c("CP default"=mean(cp$r),"CP BG fix"=mean(cpbg$r),"CP FDR fix"=mean(cpfdr$r),"CP BG and FDR fix"=mean(cpbgfdr$r),"fora"=mean(fo$r),"fgsea"=mean(fg$r))
f1 <- sapply(1:length(prec), function(i) { 2/(1/rec[i]+1/prec[i]) })

par(mfrow=c(1,3))
par(mar=c(c(9.1, 3.5, 2.1, 1.1) ))
barplot(prec,ylim=c(0,1.1),las=2,main="precision",ylab="index")
text((1:6*1.18)-0.38,prec+0.02,labels=signif(prec,3))
barplot(rec,ylim=c(0,1.1),las=2,main="recall",ylab="index")
text((1:6*1.18)-0.38,rec+0.02,labels=signif(rec,3))

barplot(f1,ylim=c(0,1.1),las=2,main="F1",ylab="index")
text((1:6*1.18)-0.38,f1+0.02,labels=signif(f1,3))
